---
title: "HW2"
author: "Zixuan_Yu"
output: html_document
---
### I.Matrix Representation of Multiple Linear Regression  
**1.Use the following 5 observations and write the simple linear regression model in matrix terms.  Then using the least squares calculations in matrix notation, compute estimates for the simple linear regression intercept and slope.**     
```{r}
y <- c(-0.1, 2.9, 6.2, 7.3, 10.7)
x <- matrix(c(1,1,1,1,1,1,3,5,7,9),nrow=5,ncol=2) ## The design matrix
x; y
```
$$Y= \begin{bmatrix}
-0.1 \\
2.9 \\
6.2 \\
7.3 \\
10.7 
\end{bmatrix}, 
X=\begin{bmatrix}
1 & 1 \\
1 & 3 \\
1 & 5 \\
1 & 7 \\
1 & 9 
\end{bmatrix}, 
\beta=\begin{bmatrix}
\beta_{1} \\
\beta_{2} 
\end{bmatrix}, 
{\epsilon}=
\begin{bmatrix}
\epsilon_1 \\
\epsilon_2 \\
\epsilon_3 \\
\epsilon_4 \\
\epsilon_5 
\end{bmatrix} $$
Model:  $Y = \beta X + \epsilon$  
Where $\epsilon_i ~ N(0, \sigma^2)$  and $Cov(\epsilon_i,\epsilon_j) = 0$
$\hat{\beta}=(X^TX)^{-1}X^TY$
So, $$ \hat{\beta} = 
(\begin{bmatrix}
1 & 1 \\
1 & 3 \\
1 & 5 \\
1 & 7 \\
1 & 9 
\end{bmatrix}^T *
\begin{bmatrix}
1 & 1 \\
1 & 3 \\
1 & 5 \\
1 & 7 \\
1 & 9 
\end{bmatrix})^{-1} 
* \begin{bmatrix}
1 & 1 \\
1 & 3 \\
1 & 5 \\
1 & 7 \\
1 & 9 
\end{bmatrix}^T *
\begin{bmatrix}
-0.1 \\
2.9 \\
6.2 \\
7.3 \\
10.7 
\end{bmatrix} =
\begin{bmatrix}
-1.1 \\
1.3 
\end{bmatrix}
$$
```{r}
solve(t(x) %*% x) %*% t(x) %*% y
```




