---
title: "hw2_1_figure"
author: "Zixuan Yu"
date: "2/25/2022"
output: html_document
---
**b. plot width **  

```{r}
## Equal Sample Size
s2 = seq(1,100, length=1000)
n1 = 1600
n2 = 1600
var_beta1 = s2/n1 + 2*s2/n2
var_beta2 = s2/n1 + 1*s2/n2
df = n1+n2-2
width1 = 2*qt(0.975, df = df )*sqrt(var_beta1)/df
width2 = 2*qt(0.975, df = df )*sqrt(var_beta2)/df
plot(s2, width1, col='lightblue')
points(s2, width2, col='pink')

```

  

```{r}
## Unequal Sample Size
s2 = seq(1,100, length=1000)
n1 = 3200
n2 = 1600
var_beta1 = s2/n1 + 2*s2/n2
var_beta2 = s2/n1 + 1*s2/n2
df = n1+n2-2
width1 = 2*qt(0.975, df = df )*sqrt(var_beta1)/df
width2 = 2*qt(0.975, df = df )*sqrt(var_beta2)/df
df1 <- data.frame(sigma2 = s2, width = c(width1,width2), model = c(rep("True model",1000), rep("model1",1000)) )
library(ggplot2)
ggplot(data=df1, aes(x=sigma2, y=width, color=model))+geom_point(size=3) +
  labs(title = "Width of 95% Confidence interval for B1") +
  theme(panel.border = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(colour = "black"),
        plot.title = element_text(size = 18, face = "bold")) 

```
  
Whether sample size are equal or not does not seems to influence the shape of the curve. The true model has larger with, and the relationship between $\sigma^2$ and width is not linear. By assuming same variance across groups, we will bias standard error done, thus getting a smaller width of the confidence interval.  A more flexible model that would allow the variance of the residuals to depend on the treatment group will more accurately reflec the reality.  



5.	OPTIONAL:  Under the Gaussian multiple linear regression framework, write the log likelihood function for the regression coefficients and residual variance in matrix terms and derive the mle’s for the regression coefficients. Derive their joint distribution, as well as the distribution of the predicted values and residuals.  